\sys: Controllable Program Regeneration

### 2025-07-28 docs/plos/draft.tex
Make wrapfigs tighter
Signed-off-by: Evangelos Lamprou <vagos@lamprou.xyz>

\begin{wrapfigure}[4]{r}{.64\columnwidth}
\begin{minted}{prolog}
fsignt(p_, S) :- fsignt(p, S).
graph(p_, I, O) :- graph(p, I, O),
                   crash(p, I).
:- crash(p_, I), graph(p_, I, _).
\end{minted}
\end{wrapfigure}
To the right is a program written in the \sys DSL to express this intent.
This query requires the regenerated program \texttt{P\_} to exhibit the same
I/O behavior on representative inputs that would cause a buffer overflow in the original program $P$.
but not on the regenerated program $P^\prime$.
\sys first arrives the set of properties to extract from the
original program, given the query:
$\{\ttt{input(P, I)}, \ttt{graph(P, I, O)}, \ttt{signature(P, S)}\}$, where
\texttt{input(P, I)} means that \sys will use extract a set of inputs from the original program
by prompting an LLM with the program's source code and asking it to generate a set of inputs that
would cause a buffer overflow in the original program;
\texttt{graph(P, I, O)} extracts the client-observable I/O behavior of the program
by executing the original program on the generated inputs and recording the outputs;
and \texttt{signature(P, S)} extracts each function's signature using a language-aware parser.

@@

\heading{An unidiomatic C function in a physics library}
The fast inverse square root routine, originally developed for the Quake III
engine~\cite{fast_inv_sqrt}, is included in the game's physics and vector math
library as an optimization for computing normalized vectors.
The implementation
exploits type punning to manipulate IEEE floats at the bit level—a technique
that relies on undefined behavior and reduces portability:
\begin{listing}
\begin{minted}{c}
float Q_rsqrt(float number) {
 long i; float x2, y;
 const float threehalfs = 1.5F;
 x2 = number * 0.5F; y  = number;
 i  = * ( long * ) &y; // evil bit level hack
 i  = 0x5f3759df - ( i >> 1 );
 y  = * ( float * ) &i;
 y  = y * ( threehalfs - ( x2 * y * y ) );
 return y; }
\end{minted}
\caption{An unidiomatic C implementation of the fast inverse square root function popularized by the game Quake III~\cite{fast_inv_sqrt}.}
\end{listing}
The query to the right uses lines of code as a coarse metric of
idiomaticity. \sys regenerates the program, minimizing this metric while
preserving observable behavior over a representative set of inputs which will be generated by an LLM instance.
\begin{wrapfigure}[2]{r}{.7\columnwidth}
  \begin{minted}{prolog}
graph(p_, I, O) :- graph(p, I, O).
#min N: len(p_, N).
  \end{minted}
\end{wrapfigure}
\sys extracts the set of properties \texttt{signature(F)} and \texttt{graph(F)}
from the original function, where \texttt{signature(F)} is the function's
signature (including its name, return type, and arguments) and
\texttt{graph(F)} captures the I/O behavior on a generated set of
floating-point inputs.
In this case, where the goal is to optimize an aspect of the program, \sys
re-prompts the model to regenerate the program until it can make no further
improvements or observes a negligible difference after two consecutive
iterations.

@@

\begin{wrapfigure}[5]{r}{.68\columnwidth}
  \begin{minted}{prolog}
db('game.db').func(f1).func(f2).
sql_trc(p_,I,T) :- sql_trc(p,I,T),
                   graph(p,I,_).
graph(call(f2,call(f1),I),I,O):-
              graph(p,I,O).
  \end{minted}
\end{wrapfigure}

@@

\section{Query Language and Synthesis}
\label{sec:synthesis}

\sys represents program characteristics as symbolic \emph{properties}, expressed as first-order predicates. Examples include:

\begin{itemize}
  \item \texttt{language(p, javascript)}: the regenerated program must be in JavaScript.
  \item \texttt{pure(p)}: the program must be side-effect free.
  \item \texttt{signature(p, S)}: the program must export a specific API.
\end{itemize}

Plugins contribute two things:
\begin{enumerate}
  \item \emph{Capability declarations}, stating which properties they can extract (e.g., \texttt{can(signature(p))}) or enforce.
  \item \emph{Logic rules}, defining relationships between properties (e.g., a translation to Haskell implies purity).
\end{enumerate}

Internally, these constraints are encoded in Answer Set Programming (ASP)~\cite{Gelfond_2000, Eiter_2009}. \sys solves for a minimal set of properties that satisfy the user’s query, then extracts them from the input program using plugin-provided analyzers.

Once the relevant properties are available, \sys assembles a structured prompt for the synthesizer. This prompt consists of:

\begin{itemize}
  \item \emph{Natural-language instructions} derived from enforced properties (e.g., "The function must be implemented in JavaScript and have no side effects.")
  \item \emph{Concrete examples}, such as input-output pairs or signature templates.
\end{itemize}

Unlike manual prompt engineering, this approach composes property constraints automatically, ensuring consistency between synthesis goals and verification criteria.

\section{Verification and Feedback}
\label{sec:verification}

After synthesis, \sys verifies whether the regenerated program satisfies all
enforced properties. Each property has a dedicated checker, contributed by the
same plugins that extract or infer it. For example:

\begin{itemize}
  \item \texttt{graph(p, I, O)}: A checker runs the synthesized code on test inputs and compares outputs.
  \item \texttt{pure(p)}: A static analyzer detects side-effecting operations.
  \item \texttt{signature(p, S)}: A parser validates exported function signatures.
\end{itemize}

If any verification fails, \sys treats this as feedback to resample or refine
the synthesis. This loop continues until all constraints are met or search is
exhausted. Partial regeneration is allowed: functions that pass verification
are retained, while others remain unchanged.
This process makes program regeneration an iterative, declarative workflow,
bridging symbolic reasoning with data-driven synthesis.

@@

\begin{table}[ht]
\centering
\caption{\textbf{Benchmark Overview}.}
\begin{tabular}{@{\extracolsep{\fill}} l r l l}
\toprule
Benchmark & N & Purpose & Sources \\
\midrule
Rosetta Code & 16 & Cross-language translation & \cite{rosettacode} \\
Short Utilities & 13 & Parity preservation & \cite{regbench2025} \\
SSCA & 3 & Malicious behavior removal & \cite{ohm2020backstabber,ev:eurosec:2022} \\
IOCCC & 2 & Deobfuscation & \cite{ioccc} \\
% Microbenchmarks & 3 & Idiomatic rewriting, modularization & This paper \\
\bottomrule
\end{tabular}
\label{tab:benchmarks}
\end{table}

@@

\heading{Fast Inverse Square Root}
The goal of this regeneration task is to transform an input 
program into a more idiomatic version that retains the same functionality.
The query for \sys is shown on the right and it uses the length of the output program as a proxy towards 
more idiomatic code.
\begin{wrapfigure}[4]{r}{0.7\columnwidth}
\begin{minted}{prolog}
graph(p_, I, O) :- graph(p, I, O).
#min N : len(p_, N)
\end{minted}
\end{wrapfigure}
The fast inverse square root function is a well-known algorithm that was
popularized by its use in the Quake III Arena source code.
It computes the inverse square root of a number using bit-level manipulation and
Newton's method for fast convergence.
The original implementation is shown in \cref{lst:fast_inv_sqrt_original}.
\Cref{lst:fast_inv_sqrt_gpt} shows the output of naively regenerating the function using \gptmodel,
which retains the original's un-idiomatic use of magic numbers and bit-level manipulation.
\Cref{lst:fast_inv_sqrt_sys} has the output of \sys, which instead uses the standard library's
\ttt{sqrt} function to compute the inverse square root.

>>>>

\begin{wrapfigure}[4]{r}{.53\columnwidth}
\begin{minted}{prolog}
fsignt(p_,S):-fsignt(p,S).
graph(p_,I,O):-graph(p,I,O),
               crash(p,I).
:-crash(p_,I),graph(p_,I,_).
\end{minted}
\end{wrapfigure}
To the right is a program written in the \sys DSL to express this intent.
This query requires the regenerated program \texttt{P\_} to exhibit the same
I/O behavior on representative inputs that would cause a buffer overflow in the original program $P$.
but not on the regenerated program $P^\prime$.
\sys first arrives the set of properties to extract from the
original program, given the query:
$\{\ttt{input(P, I)}, \ttt{graph(P, I, O)}, \ttt{signature(P, S)}\}$, where
\texttt{input(P, I)} means that \sys will use extract a set of inputs from the original program
by prompting an LLM with the program's source code and asking it to generate a set of inputs that
would cause a buffer overflow in the original program;
\texttt{graph(P, I, O)} extracts the client-observable I/O behavior of the program
by executing the original program on the generated inputs and recording the outputs;
and \texttt{signature(P, S)} extracts each function's signature using a language-aware parser.

@@

\heading{An unidiomatic C function in a physics library}
The fast inverse square root routine, originally developed for the Quake III
engine~\cite{fast_inv_sqrt}, is included in the game's physics and vector math
library as an optimization for computing normalized vectors.
The implementation
exploits type punning to manipulate IEEE floats at the bit level—a technique
that relies on undefined behavior and reduces portability:
\begin{listing}
\begin{minted}{c}
float Q_rsqrt(float number) {
 long i; float x2, y;
 const float threehalfs = 1.5F;
 x2 = number * 0.5F; y  = number;
 i  = * ( long * ) &y; // evil bit level hack
 i  = 0x5f3759df - ( i >> 1 );
 y  = * ( float * ) &i;
 y  = y * ( threehalfs - ( x2 * y * y ) );
 return y; }
\end{minted}
\caption{An unidiomatic C implementation of the fast inverse square root function popularized by the game Quake III~\cite{fast_inv_sqrt}.}
\end{listing}
The query to the right uses lines of code as a coarse metric of
idiomaticity. \sys regenerates the program, minimizing this metric while
preserving observable behavior over a representative set of inputs which will be generated by an LLM instance.
\begin{wrapfigure}[2]{r}{.55\columnwidth}
\begin{minted}{prolog}
graph(p_,I,O):-graph(p,I,O).
#min N: len(p_, N).
\end{minted}
\end{wrapfigure}
\sys extracts the set of properties \texttt{signature(F)} and \texttt{graph(F)}
from the original function, where \texttt{signature(F)} is the function's
signature (including its name, return type, and arguments) and
\texttt{graph(F)} captures the I/O behavior on a generated set of
floating-point inputs.
In this case, where the goal is to optimize an aspect of the program, \sys
re-prompts the model to regenerate the program until it can make no further
improvements or observes a negligible difference after two consecutive
iterations.

@@

\begin{wrapfigure}[5]{r}{.60\columnwidth}
  \begin{minted}{prolog}
db('game.db').func(f1).func(f2).
sql_trc(p_,I,T):-sql_trc(p,I,T),
                  graph(p,I,_).
graph(call(f2,call(f1),I),I,O):-
                  graph(p,I,O).
  \end{minted}
\end{wrapfigure}

@@

@@

\begin{table}[ht]
\centering
\caption{\textbf{Benchmark Overview}.}
\begin{tabular}{lrll}
\toprule
Benchmark & N & Purpose & Sources \\
\midrule
Rosetta Code & 16 & Cross-language translation & \cite{rosettacode} \\
Short Utilities & 13 & Parity preservation & \cite{regbench2025} \\
SSCA & 3 & Malicious behavior removal & \cite{ohm2020backstabber,ev:eurosec:2022} \\
IOCCC & 2 & Deobfuscation & \cite{ioccc} \\
% Microbenchmarks & 3 & Idiomatic rewriting, modularization & This paper \\
\bottomrule
\end{tabular}
\label{tab:benchmarks}
\end{table}

@@

\heading{Fast Inverse Square Root}
The goal of this regeneration task is to transform an input 
program into a more idiomatic version that retains the same functionality.
The query for \sys is shown on the right and it uses the length of the output program as a proxy towards 
more idiomatic code.
\begin{wrapfigure}[4]{r}{0.65\columnwidth}
\begin{minted}{prolog}
graph(p_, I, O) :- graph(p, I, O).
#min N : len(p_, N)
\end{minted}
\end{wrapfigure}
The fast inverse square root function is a well-known algorithm that was
popularized by its use in the Quake III Arena source code.
It computes the inverse square root of a number using bit-level manipulation and
Newton's method for fast convergence.
The original implementation is shown in \cref{lst:fast_inv_sqrt_original}.
\Cref{lst:fast_inv_sqrt_gpt} shows the output of naively regenerating the function using \gptmodel,
which retains the original's un-idiomatic use of magic numbers and bit-level manipulation.
\Cref{lst:fast_inv_sqrt_sys} has the output of \sys, which instead uses the standard library's
\ttt{sqrt} function to compute the inverse square root.
